Introducing the \NAME Framework\\ for Efficient Intrusion Detection\\ in the Internet of Things

### 2014-10-16 paper.tex
UK -> US

Supporting multiple intrusion detection (ID) algorithms imposes significant
overhead on Internet of Things (IoT) devices. Fusion of the underlying
algorithms can alleviate this, yet, proves to be time-consuming, repetitive and
error-prone. To address this problem we introduce \NAME, a framework for the
development of efficient intrusion detection systems (IDS). It consists of a
domain specific language (DSL) that allows formally describing ID algorithms on
a functional level. A code generator then fuses these algorithms and produces
more optimally organised source code. This way, \NAME addresses the
heterogeneity of both the IoT and ID, as well as the limited resources of IoT
devices. A side-by-side comparison shows that generated code reduces memory
footprint, message passing overhead and execution time in comparison to a
straightforward sequential implementation of the same algorithms.

@@

In mobile ad hoc networks (MANET), which make up an important part of the IoT,
it is not possible to have this single point in the network that can oversee
all traffic \cite{zhang2000intrusion, mishra2004intrusion}. If the outer
security perimeter diminishes, every device has to implement its own lines of
defence. Making the IDS a local service on each device requires resources,
which are another important constraint: IoT devices typically have limited
batteries, little processing power and memory is restricted to the bare minimum.

@@

Current developments in ID focus on programming frameworks to structure the
implementation of different algorithms \cite{valero2012di} and offer the
required basic functional components to implement actual algorithms
\cite{krontiris2008lidea}, but don't offer a way to optimise the usage of
resources. They also don't support systematic reuse to create different
configurations for different nodes or evolution in time.

@@

The second contribution introduces functional code fusion (FCF) as a generation
paradigm. FCF identifies common data and functions, and organises code in such
a way that redundant iterations, tests and computations are eliminated. Our
results show substantially improved memory usage, execution time and usage of
the wireless radio, reducing energy consumption.

@@

The remainder of this paper proceeds as follows: section \ref{background}
details some of the ideas behind FCF and identifies the different aspects of
the problem space. Section \ref{design} describes the design we applied to
construct the framework, its domain specific language (DSL) and code generator.
Section \ref{evaluation} evaluates an implementation and application of the
framework. Section \ref{related} explores related work in the field of DSLs for
ID and code generation for the IoT. Finally, section \ref{conclusion}
summarises our findings, draws conclusions and identifies topics for future
work.

@@

\begin{figure}[H]
\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
\begin{algorithmic}[1]
  \Require{nodes, global storage for information about nodes}
  \Function{process\_message}{$msg$}
    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
     \State \dots \Comment{analyse byte-sequences}
    \EndFor
    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
  \EndFunction
  \State
  \Function{do\_housekeeping}{}
    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
      \If{$node > \dots$} \Comment{validate recorded value}
        \State \dots \Comment{take actions}
        \State \Call{send}{$node$, ``info''} \Comment{e.g. communicate} \label{alg:id-algo-pattern-send2}
      \EndIf
    \EndFor
  \EndFunction
  \State
\end{algorithmic}
\end{figure}

@@

Starting from scratch allows optimising the implementation of an IDS, but
doesn't scale. Reusing existing implementations also doesn't help, due to the
far from optimal sequential execution of the algorithms and the redundant
storage of common data. Software libraries can solve many of the issues, but
only cover parts of the entire spectrum, due to their tie-in with languages and
frameworks.

@@

A semantic model \cite{fowler2010domain}, tied closely to the DSL itself, can
fulfil this role. Semantic transformations can now be applied to create a
code-oriented model, comparable to the idea of CodeDom. From there on,
syntactic transformations an emit actual source code. Figure
\ref{fig:code-generation} shows the overall procedure.

@@

The DSL is parsed and imported into a semantic model. Based on the minimal type
information present in the model, all remaining types are inferred. Through a
series of semantic transformations, the model evolves from the functional to a
more technical one, a code model. This code model can be considered as a rich
AST with an extensive set of language constructs. Syntactic transformations now
lower this level of expressiveness according to the desired platform and
language. In the end the code model can be directly emitted to the target
language.

@@

The semantic model is the core of the solution. It offers a way to describe ID
algorithms in a way that allows semantic reinterpretation. This way, the
functionality of different algorithms can be reorganised to allow for more
optimal execution and memory usage. Hence the name \NAME of the framework:
intrusion detection functionality organisation optimisation.

@@

From a functional point of view, the generated code lives up to the
expectations: both the number of frames as the bytes sent benefit from well
organised code and reuse of a common software library.

@@

Without any effort to optimise the software library, nor leveraging knowledge
about the specific algorithms, this very basic generated code shows that the
proposed principles are not only theoretically feasible but actually result in
interesting improvements.

>>>>

Supporting multiple intrusion detection (ID) algorithms imposes significant
overhead on Internet of Things (IoT) devices. Fusion of the underlying
algorithms can alleviate this, yet, proves to be time-consuming, repetitive and
error-prone. To address this problem we introduce \NAME, a framework for the
development of efficient intrusion detection systems (IDS). It consists of a
domain specific language (DSL) that allows formally describing ID algorithms on
a functional level. A code generator then fuses these algorithms and produces
more optimally organized source code. This way, \NAME addresses the
heterogeneity of both the IoT and ID, as well as the limited resources of IoT
devices. A side-by-side comparison shows that generated code reduces memory
footprint, message passing overhead and execution time in comparison to a
straightforward sequential implementation of the same algorithms.

@@

In mobile ad hoc networks (MANET), which make up an important part of the IoT,
it is not possible to have this single point in the network that can oversee
all traffic \cite{zhang2000intrusion, mishra2004intrusion}. If the outer
security perimeter diminishes, every device has to implement its own lines of
defense. Making the IDS a local service on each device requires resources,
which are another important constraint: IoT devices typically have limited
batteries, little processing power and memory is restricted to the bare minimum.

@@

Current developments in ID focus on programming frameworks to structure the
implementation of different algorithms \cite{valero2012di} and offer the
required basic functional components to implement actual algorithms
\cite{krontiris2008lidea}, but don't offer a way to optimize the usage of
resources. They also don't support systematic reuse to create different
configurations for different nodes or evolution in time.

@@

The second contribution introduces functional code fusion (FCF) as a generation
paradigm. FCF identifies common data and functions, and organizes code in such
a way that redundant iterations, tests and computations are eliminated. Our
results show substantially improved memory usage, execution time and usage of
the wireless radio, reducing energy consumption.

@@

The remainder of this paper proceeds as follows: section \ref{background}
details some of the ideas behind FCF and identifies the different aspects of
the problem space. Section \ref{design} describes the design we applied to
construct the framework, its domain specific language (DSL) and code generator.
Section \ref{evaluation} evaluates an implementation and application of the
framework. Section \ref{related} explores related work in the field of DSLs for
ID and code generation for the IoT. Finally, section \ref{conclusion}
summarizes our findings, draws conclusions and identifies topics for future
work.

@@

\begin{figure}[H]
\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
\begin{algorithmic}[1]
  \Require{nodes, global storage for information about nodes}
  \Function{process\_message}{$msg$}
    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
     \State \dots \Comment{analyze byte-sequences}
    \EndFor
    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
  \EndFunction
  \State
  \Function{do\_housekeeping}{}
    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
      \If{$node > \dots$} \Comment{validate recorded value}
        \State \dots \Comment{take actions}
        \State \Call{send}{$node$, ``info''} \Comment{e.g. communicate} \label{alg:id-algo-pattern-send2}
      \EndIf
    \EndFor
  \EndFunction
  \State
\end{algorithmic}
\end{figure}

@@

Starting from scratch allows optimizing the implementation of an IDS, but
doesn't scale. Reusing existing implementations also doesn't help, due to the
far from optimal sequential execution of the algorithms and the redundant
storage of common data. Software libraries can solve many of the issues, but
only cover parts of the entire spectrum, due to their tie-in with languages and
frameworks.

@@

A semantic model \cite{fowler2010domain}, tied closely to the DSL itself, can
fulfill this role. Semantic transformations can now be applied to create a
code-oriented model, comparable to the idea of CodeDom. From there on,
syntactic transformations an emit actual source code. Figure
\ref{fig:code-generation} shows the overall procedure.

@@

The DSL is parsed and imported into a semantic model. Based on the minimal type
information present in the model, all remaining types are inferred. Through a
series of semantic transformations, the model evolves from the functional to a
more technical one, a code model. This code model can be considered as a rich
abstract syntax tree (AST) with an extensive set of language constructs.
Syntactic transformations now lower this level of expressiveness according to
the desired platform and language. In the end the code model can be directly
emitted to the target language.

@@

The semantic model is the core of the solution. It offers a way to describe ID
algorithms in a way that allows semantic reinterpretation. This way, the
functionality of different algorithms can be reorganized to allow for more
optimal execution and memory usage. Hence the name \NAME of the framework:
intrusion detection functionality organization optimization.

@@

From a functional point of view, the generated code lives up to the
expectations: both the number of frames as the bytes sent benefit from well
organized code and reuse of a common software library.

@@

Without any effort to optimize the software library, nor leveraging knowledge
about the specific algorithms, this very basic generated code shows that the
proposed principles are not only theoretically feasible but actually result in
interesting improvements.
